Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-18 paper/paper.tex
Updates from Overleaf

The most studied version of the above problems is finding a maximum/minimum set satisfying the contraints and the corresponding set size,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. it is unlikely to be solved in a polynomial time.
\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied \textit{solution space property} computation which can be harder, e.g. counting independent sets of a general graph is in \#P-complete~\cite{Vadhan2001}, but is crucial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

We propose to view the computation of all these properties from an abstract algebra perspective and use the unified framework of
generic tensor-network to solve these seemingly unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy.
It is composed of a set of tensors and a set of labels for specifying the contraction of tensors,
where the tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is 2 power the treewidth of its line graph~\cite{Markov2008}.
\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best sets with the maximum size,
this time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
These two algorithms are related but are not equivalent.
Dynamic programming is more general purposed, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

To avoid the discussion being unnecessarily diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
\xpurple{More details on which combinatorial optimization problems this
applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting}, \Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

@@

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}

\purple{I think the following paragraph could be clarified. When you read the definition
of generic programming, it's not clear what ``in the most general setting'' means. My
understanding is it just means that the program will work regardless of the input data
type - it would be good to say this explicitly. }
\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
\purple{I also think the way it is written takes a while
to understand what dynamically versus statically typed languages are. The definition,
“When these languages “see” a new input type, the compiler can recompile the generic
program for the new type.” comes rather late in the paragraph.}
\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
lost on someone only with a physics background.}
This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
To understand the first aspect on generality, suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their numbers
   \purple{``and their counting'' is gramatically confusing, can we say independence polynomial or degeneracy or something?}\blue{Since you mentioned degeneracy is bad, what about ``numbers''?}} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

It is in general not computationally efficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products, which scales exponentially as the number of vertices.
\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor contraction.
A fully optimized contraction order can reduce the time complexity significantly, while requiring a minimum space complexity $O(2^{\text{tw}(G)})$~\cite{Markov2008} to store intermediate contraction results.
\purple{Perhaps just say time and space
complexity? Order of magnitude doesn't make sense to me in terms of scaling. Also,
what is the time complexity of finding a good contraction order? Is there a known
average-case complexity? This seems important.}\blue{see added texts below.}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
In practise, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order in typically a few minutes with a heuristic contraction order finding algorithm~\cite{Kourtis2019,Kalachev2021}.
For large scale applications,
it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
enumerating certain degrees of freedom in an external loop so that the space complexity of contracting a smaller tensor network inside the loop can be smaller.

@@

This algebra is a commutative semiring because we define the equivalence of two expression trees by comparing their expanded (using \Eq{eq:set}) forms, rather than their storages.
Although one is probably unable to enumerate the configurations represented by this tree due to the extremely large configuration space size,
but he can directly sample configurations from this tree.
This is because once we have this expression tree representation of the configuration space,
either generate by using this algebra directly or serve as the coefficients of (truncated) polynomials,
one can compute the number of configurations in each branch rigorously and efficiently.
\purple{To clarify - is the expression tree now the coefficients
on the polynomial? I am a little confused on how we would, for instance, use this to
compute the MISs}\blue{added one more descriptive sentence, does it resolve your concern?}
To generate samples, one just descends this expression tree to left or right sibling with probability decided by the size of each branch while doing the computation specified by the first field.
\xpurple{How do we compute the size of each
branch?}
\begin{example}
Let us consider the following expression tree 
\begin{equation*}
    (\texttt{SUM}, /, (\texttt{PROD}, /, A, B), (\texttt{SUM}, /, C, D)),
\end{equation*}
where siblings $A, B, C$ and $D$ can be any four types of nodes.
It can be represented diagramatically as the following.

@@

A theoretically interesting number that can be computed using the number of independent sets is the entropy constant for the hardcore lattice gases on these graphs, which can tell us thermal dynamic properties of these lattices at the high temperature limit.
\purple{Why is it useful and
what does it tell us about the physical system?}\blue{It is probably useless, so: useful -> theoretically interesting}
For the square grid graphs, it is called the \textit{hard square entropy constant} (\href{https://oeis.org/A085850}{OEIS A085850}), which is defined as $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$, where $F(L,L)$ is the number of independent sets of a given lattice dimensions $L \times L$. This quantity arises in statistical mechanics of hard-square lattice gases~\cite{Baxter1980, Pearce1988} and is used to understand phase transitions for these systems. This entropy constant is not known to have an exact representation, but it is accurately known in many digits. Similarly, we can define entropy constants for other lattice gases. In \Fig{fig:hardsquare}, we look at how $F(L,L)^{1/\lfloor pL^2\rceil}$ scales as a function of the grid size $L$ for all types of graphs shown in Figure~\ref{fig:lattices}. Our results match the known results for the non-disordered square grid and King's graphs. For disordered square grid and King's graphs with a filling factor $p=0.8$, we randomly sample 1000 graph instances. To our knowledge, the entropy constants for these disordered graphs have not been studied before. Interestingly, the variations due to different random instances are negligible for this entropy quantity.

@@

\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/exp_configurations.pdf}
    \caption{A visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem.
    Each vertex represents an independent set, with edges between independent sets with pairwise Hamming distance $\leq 2.$ \blue{can be connected by a swap operation?}}
    \label{fig:exp_configuratoins}
\end{figure}
The ability to enumerate configurations can also be used to understand the distribution of outputs from algorithms for solving the Maximum Independent Set problem.
This can give detailed insights to performance of the algorithm, such as if the algorithm is likely to return independent sets from certain clusters of local or global minima in the solution space.
As an example, we show in Figure~\ref{fig:exp_configuratoins} how our algorithm can be used to visualize experimental outputs of a quantum algorithm for solving the Maximum Independent Set problem using a programmable Rydberg atom array quantum computer~\cite{Ebadi2022}.
The structure of the solution space for a King's graph with $39$ vertices and filling factor $\rho=0.8$ can be visualized on a graph where each vertex represents a large independent set, with edges between independent sets with a small pairwise Hamming distance $\leq 2.$
The algorithm either finds the largest independent set with size $\alpha(G)$ with high probability, or returns a suboptimal independent set, usually with size $\alpha(G)-1.$
The algorithm does not appear to return local minima with large Hamming distance from the MISs as suggested by~\cite{altshuler2010},
but samples from local minima with a wide range of Hamming distances from the MISs.
Consistent with Sec.~\ref{sec:overlap-gap}, there are no disconnected clusters in the solution space for this particular King's graph.

>>>>

The most studied aspect of the above problems is finding an optimal solution, 
%The most studied version of the above problems is finding an optimal solution maximum/minimum set satisfying the contraints and the corresponding set size,
, which is typically 
%while many of them belong to the hardness category
NP-hard~\cite{Hastad1996}, i.e. it is unlikely to be solved in a polynomial time.
%\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
%\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied problem of computing \textit{solution space properties}  which can be much harder, e.g. counting independent sets of a general graph is \#P-complete~\cite{Vadhan2001}, but is crucial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

We propose to view the computation of all these properties from an abstract algebra perspective and use the unified framework of
generic tensor-network to solve these seemingly unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy.
It is composed of a set of tensors and a set of labels for specifying the contraction of tensors,
where the tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is
two to the power of the treewidth of its line graph~\cite{Markov2008}.
%\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time using heuristic (inexact) methods.
%\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best sets with the maximum size,
the time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
However, the tensor network has a richer algebraic structure, which crucially enables the computation of solution space properties beyond finding the solution, which is not possible with dynamic programming.
%These two algorithms are related but are not equivalent.
%Dynamic programming is more general purpose, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
%\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

To avoid the discussion being unnecessarily diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
%\xpurple{More details on which combinatorial optimization problems this applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting}, \Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

@@

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}
\purple{efficiency}
By in the most general setting, we mean that a single program should work correctly for different input data types. For example, 
%\purple{I think the following paragraph could be clarified. When you read the definition of generic programming, it's not clear what ``in the most general setting'' means. My understanding is it just means that the program will work regardless of the input data type - it would be good to say this explicitly. }
%\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
%\purple{I also think the way it is written takes a while to understand what dynamically versus statically typed languages are. The definition, “When these languages “see” a new input type, the compiler can recompile the generic program for the new type.” comes rather late in the paragraph.}
%\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
%lost on someone only with a physics background.}
%This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
%To understand the first aspect on generality,
suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

@@

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their degeneracy} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

It is in general not computationally efficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products. %, which scales exponentially as the number of vertices.
%\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor contraction.
A fully optimized contraction order can reduce the time complexity significantly, while requiring a minimum space complexity $O(2^{\text{tw}(G)})$~\cite{Markov2008} to store intermediate contraction results.
In practice, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order in typically a few minutes with a heuristic contraction order finding algorithm~\cite{Kourtis2019,Kalachev2021}. The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
For large scale applications,
it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
enumerating certain degrees of freedom in an external loop so that the space complexity of contracting a smaller tensor network inside the loop can be smaller.

@@

This algebra is a commutative semiring because we define the equivalence of two expression trees by comparing their expanded (using \Eq{eq:set}) forms, rather than their storages.
Although one is probably unable to enumerate the configurations represented by this tree due to the extremely large configuration space size,
but he can directly sample configurations from this tree.
This is because once we have this expression tree representation of the configuration space,
either generate by using this algebra directly or serve as the coefficients of (truncated) polynomials,
one can compute the number of configurations in each branch rigorously and efficiently.
%\purple{To clarify - is the expression tree now the coefficients
%on the polynomial? I am a little confused on how we would, for instance, use this to
%compute the MISs}\blue{added one more descriptive sentence, does it resolve your concern?}
To generate samples, one just descends this expression tree to left or right sibling with probability decided by the size of each branch while doing the computation specified by the first field.
\xpurple{How do we compute the size of each
branch?}
\begin{example}
Let us consider the following expression tree 
\begin{equation*}
    (\texttt{SUM}, /, (\texttt{PROD}, /, A, B), (\texttt{SUM}, /, C, D)),
\end{equation*}
where siblings $A, B, C$ and $D$ can be any four types of nodes.
It can be represented diagramatically as the following.

@@

A theoretically interesting number that can be computed using the number of independent sets is the entropy constant for the hardcore lattice gases on these graphs, which can tell us thermal dynamic properties of these lattices at the high temperature limit.
For the square grid graphs, it is called the \textit{hard square entropy constant} (\href{https://oeis.org/A085850}{OEIS A085850}), which is defined as $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$, where $F(L,L)$ is the number of independent sets of a given lattice dimensions $L \times L$. This quantity arises in statistical mechanics of hard-square lattice gases~\cite{Baxter1980, Pearce1988} and is used to understand phase transitions for these systems. This entropy constant is not known to have an exact representation, but it is accurately known in many digits. Similarly, we can define entropy constants for other lattice gases. In \Fig{fig:hardsquare}, we look at how $F(L,L)^{1/\lfloor pL^2\rceil}$ scales as a function of the grid size $L$ for all types of graphs shown in Figure~\ref{fig:lattices}. Our results match the known results for the non-disordered square grid and King's graphs. For disordered square grid and King's graphs with a filling factor $p=0.8$, we randomly sample 1000 graph instances. To our knowledge, the entropy constants for these disordered graphs have not been studied before. Interestingly, the variations due to different random instances are negligible for this entropy quantity.

@@

\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/exp_configurations.pdf}
    \caption{A visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem.
    Each vertex represents an independent set, with edges between independent sets which differ by a swap operation or a vertex addition.}
    \label{fig:exp_configuratoins}
\end{figure}
The ability to enumerate configurations can also be used to understand the distribution of outputs from algorithms for solving the Maximum Independent Set problem.
This can give detailed insights to performance of the algorithm, such as if the algorithm is likely to return independent sets from certain clusters of local or global minima in the solution space.
As an example, we show in Figure~\ref{fig:exp_configuratoins} how our algorithm can be used to visualize experimental outputs of a quantum algorithm for solving the Maximum Independent Set problem using a programmable Rydberg atom array quantum computer~\cite{Ebadi2022}.
The structure of the solution space for a King's graph with $39$ vertices and filling factor $\rho=0.8$ can be visualized on a graph where each vertex represents a large independent set, with edges between independent sets which differ by a swap operation or a vertex addition.
The algorithm either finds the largest independent set with size $\alpha(G)$ with high probability, or returns a suboptimal independent set, usually with size $\alpha(G)-1.$
The algorithm does not appear to return local minima with large Hamming distance from the MISs as suggested by~\cite{altshuler2010},
but samples from local minima with a wide range of Hamming distances from the MISs.
Consistent with Sec.~\ref{sec:overlap-gap}, there are no disconnected clusters in the solution space for this particular King's graph.
